---
title: "Introduction to predictsr"
output: html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to predictsr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
cache: true
---

The `{predictsr}` package is used to access the PREDICTS database from within `R`. It provides a set of functions that access the [NHM Data Portal](https://data.nhm.ac.uk/) to download the latest versions of the PREDICTS database and some related metadata easily, from within R. The [PREDICTS](https://www.nhm.ac.uk/our-science/research/projects/predicts/science.html) database comprises over 4 million measurements of species at sites across the world. The data are mostly collected from arthropods (mainly insects), and we cover about 2% of the species known by science. All major terrestrial plant, animal, and fungal groups are covered.

There have been 2 public releases of PREDICTS data. The first was in [2016](https://data.nhm.ac.uk/dataset/the-2016-release-of-the-predicts-database-v1-1) and the second was in [2022](https://data.nhm.ac.uk/dataset/release-of-data-added-to-the-predicts-database-november-2022); each consisted of 3 million and 1 million records, respectively. We include both, as a single dataset, in this package.

To get started, let's load in the data into `R` and poke around. To do so you will use the `GetPredictsData` function, which pulls in the data from the data portal.

I'll repeat the default options below, but you can specify either a `"data.frame"` or `"tibble"` as the format, and take either 2016, 2022, or both as the extracts to fetch. The default is to return a dataframe, and read in both extracts and concatenate into one:

```{r setup}
predicts <- predictsr::GetPredictsData(fmt = "data.frame", extract = c(2016, 2022))
str(predicts)
```

So we can see that there are over 4 million records in the combined PREDICTS extracts. Let's look at a set of summary statistics for the database:

```{r}
print(glue::glue(
  "This database has {length(unique(predicts$SS))} studies across ",
  "{length(unique(predicts$SSBS))} sites, in ",
  "{length(unique(predicts$Country))} countries, and with ",
  "{length(unique(predicts$Species))} species."
))
```

So over 30,000 sites, 101 countries, and 12,892 species in this dataframe! There are a couple of important columns that should be noted. The `SS` column is generally what we use to identify studies when dealing with PREDICTS data. This is the concantenation of the `Source_ID`, and the `Study_number` columns. Another important identifier is the `SBBS`, the concatenation of `Source_ID`, `Study_number`, `Block` and `Site_number`, which is what we use to identify single sites in the database.

Let's also check the ranges of sample collection in the database:

```{r}
print(glue::glue(
  "Earliest sample collection (midpoint): {min(predicts$Sample_midpoint)}, ",
  "latest sample collection (midpoint): {max(predicts$Sample_midpoint)}"
))
```

## Accessing site-level summaries

We also include access to the site-level summaries from the full release. To get *these data* you will need to use the `GetSitelevelSummaries` function. It has the same arguments as the `GetPredictsData` function, so the function call is very similar to pull in the summaries for the same data as above:

```{r}
summaries <- predictsr::GetSitelevelSummaries(fmt = "data.frame", extract = c(2016, 2022))
```

## Accessing descriptions of the columns in PREDICTS

We have 67 (!) columns in the PREDICTS database, and within the NHM Data portal releases, we have included a description of the data that is used in each of these columns. You can access this via the `GetColumnDescriptions` function:

```{r}
descriptions <- predictsr::GetColumnDescriptions(fmt = "data.frame")
str(descriptions)
```

### Database columns and notes

The columns are described as follows:

```{r, echo = FALSE, results = 'asis'}
for (i in seq_along(descriptions$Column)) {
  cat(
    paste0("`", descriptions$Column[i], "` (`", descriptions$Type[i], "`)  \n")
  )
  notes <- descriptions$Notes[i] |>
    (\(s) gsub("\\*", "-", s))() |>
    (\(s) gsub("\n", "  \n", s))()

  if (notes == " ") {
    notes <- "As title."
  }

  cat(notes)
  cat("  \n  \n")
}
```


